Does personal relevance attenuate citizens' persuasion by government branding? Evidence from a survey experiment on air-pollution policy

### 2019-07-22 paper3_revision_text.Rmd
Add files via upload

---
title: "Does personal relevance attenuate citizens' persuasion by government branding? Evidence from a survey experiment on air-pollution policy"
author: '**Saar Alon-Barkat**'
date: " "
always_allow_html: yes
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
urlcolor: blue
---

@@

This study contributes to the growing Public-administration research on government branding and its potency to influence citizens' perceptions, and their inclination to trust organizations' policy plans. I explored whether persuasion through branding and symbolic communication can also ocure in policy matters that have significant, clear implications for citizens' well-being. I examined the hypothesis, derived from social psychology theory, that persuasion via symbols is attenuated the more citizens perceive the communication as personally relevant. I put this hypothesis to rigorous empirical testing using a randomized survey experiment focused on air-pollution policy, thus exploiting a natural variation in the perceived personal relevance between citizens residing in a polluted area and outside it. In addition to utilizing these natural differences, I also experimentally manipulated the personal relevance of the policy. To the best of my knowledge, this is the first study to explore the moderating role of perceived personal relevance on citizens' responses to government symbolic communications or branding.

Contrary to my expectations, and those of extant research in social psychology and marketing, I did not find that higher levels of perceived personal relevance attenuate the effect of symbols. While subjects residing in the polluted Haifa Bay reported in the survey that they perceive the governmental policy for reducing air-pollution in that area as highly relevant for their lives, they were similarly, and perhaps even more affected by the appearance of familiar brand elements of the environmental protection ministry in the communication of that policy plan. Congruently, the study indicates that Haifa Bay subjects did not pay more attention to the logical quality of the policy plan. Furthermore, the results tend to indicate that priming Haifa Bay subjects to think about the personal relevance of the air-pollution did not significantly attenuate their susceptibility to persuasion by the symbols.

How can we explain the finding that the positive effect of the real symbols was stronger among Haifa Bay residents? Given that this pattern is also evident for the recycling policy, it is unlikely that it is linked to differences in perceived personal relevance. Rather, I suggest, tentatively, that it might simply be due to the lower baseline of Haifa Bay subjects' trust in EPM policies. Accordingly, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects), indicating a ceiling effect. Stated otherwise, symbols may be more effective in reducing citizens' distrust, as opposed to increasing their trust.[^footnote_ceiling_effect] A related explanation is that the symbols distracted the subjects and decreased their inclination to think about the topic. As a result, they caused Haifa Bay subjects to disregard their prior beliefs about EPM and its failure regarding air-pollution.[^footnote_mediation_elaboration]

[^footnote_ceiling_effect]: This tentative explanation supported by the comparison of the percentages of subjects who reported very low levels of trust in the policies. This comparison consistently shows a pattern of decreased marginal reduction. In the Haifa Bay, the percentage of those who reported a trust score of less than 2 in the air-pollution policy decreased from `r ((filter(SVIVA2_01_haifa,SYMBOL==0,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==0)%>%nrow())*100)%>%round(1)`% in the no symbols group to `r ((filter(SVIVA2_01_haifa,SYMBOL==2,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==2)%>%nrow())*100)%>%round(1)`% in the real symbols. Among the center subjects, only `r ((filter(SVIVA2_01_center,SYMBOL==0,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==0)%>%nrow())*100)%>%round(1)`% of the no symbols group reported a trust score lower than 2 in the first place, and this percentage dropped to `r ((filter(SVIVA2_01_center,SYMBOL==2,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==2)%>%nrow())*100)%>%round(1)`% among the real symbols group. With regard to the recycling policy, the percentage decreased from `r ((filter(SVIVA2_01_haifa,SYMBOL==0,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==0)%>%nrow())*100)%>%round(1)`% to `r ((filter(SVIVA2_01_haifa,SYMBOL==2,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==2)%>%nrow())*100)%>%round(1)`% in the Haifa Bay and from `r ((filter(SVIVA2_01_center,SYMBOL==0,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==0)%>%nrow())*100)%>%round(1)`% to `r ((filter(SVIVA2_01_center,SYMBOL==2,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==2)%>%nrow())*100)%>%round(1)`% in the center.

[^footnote_mediation_elaboration]: This explanation is further supported by the survey data. The real symbols had a negative effect on the elaboration of these policies,  as measured by reaction time and memory test. The decrease in elaboration mediates the effect of symbols on trust.

The limitations of this study should be acknowledged. The study focuses, empirically, on one local policy case. Future studies may put these findings to further test by testing cases of air-pollution policy in other countries, as well as by exploring additional cases of public organizations and policy areas. Additionally, as already acknowledged in the paper, I cannot entirely rule out the possibility that the differences in personal relevance between the two geographic areas are confounded with other differences associated with people's beliefs about the subject. Future studies may put these findings to a more rigorous test through natural experimental or quasi-experimental designs that take advantage of random allocation of citizens to varying levels of personal relevance. Finally, whereas this study includes an experimental manipulation for perceived personal relevance, I acknowledge the limitations of that test, namely its statistical power and the fact that the data does not confirm the its effectiveness in altering the moderating variable. Therefore, future studies may use different, possibly stronger experimental manipulations for testing that variable.

@@

The results of this study, if generalizable, have significant implications. They suggest that citizens are susceptible to persuasion by government branding and symbolic communication, even with regard to salient, consested policy issues that have significant consequences for their own lives. Altogether, this study highlights the role played by government branding in shaping citizens view of governments. It suggests that the boundaries of persuasion via branding are much wider than expected. From the practical point of view of government organizations, this entails that they can perhaps overcome citizens automatic skepticism and distrust by investing in strategic symbolic communications. Nonetheless, as discussed in the introduction, employing a broader perspective of democratic theory, these conclusions are disturbing since they they buttress the concern that government organizations will deliberately exploit branding to cause citizens to view them more favorably regardless (or even in contrast) to their actual capacities and achievements. For good or for ill, citizens are affected by the emotional connections and associations of symbols.

>>>>

---
title: "Does personal relevance attenuate citizens' persuasion by government branding? Evidence from a survey experiment on air-pollution policy"
author: '**Saar Alon-Barkat**'
date: " "
always_allow_html: yes
output:
  word_document: default
  pdf_document:
    toc: yes
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
link-citations: yes
bibliography: phd_paper_2.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
urlcolor: blue
---

@@

This study contributes to the growing public-administration research on government branding and its potency to influence and possibly manipulate citizens' perceptions of government organizations and their policies and services. I explored whether persuasion through branding and symbolic communication can also ocure in policy matters that have significant, clear implications for citizens' well-being. I examined the hypothesis, derived from social psychology theory, that persuasion via symbols is attenuated the more citizens perceive the communication as personally relevant. I put this hypothesis to rigorous empirical testing using a randomized survey experiment focused on air-pollution policy, utilizing a natural variation in the perceived personal relevance between citizens residing in a polluted area and others, residing elsewhere. To the best of my knowledge, this is the first study to explore, theoretically and empirically, the moderating role of perceived personal relevance on citizens' responses to government branding, and more generally to government public communications.

Contrary to my expectations, and those of extant research in social psychology and marketing, I did not find that higher levels of perceived personal relevance attenuate the effect of symbols. While subjects residing in the polluted Haifa Bay reported in the survey that they perceive the governmental policy for reducing air-pollution in that area as highly relevant for their lives, they were similarly, and perhaps even more affected by the appearance of familiar brand elements of the environmental protection ministry in the communication of that policy plan. Congruently, they did not pay more attention to the substantive content of the policy plan. Furthermore, the results indicate that even priming Haifa Bay residents to think about the personal relevance of the air-pollution did not significantly attenuate their susceptibility to persuasion by the symbols.

@@

How can we explain the finding that the positive effect of the real symbols was stronger to some extent among Haifa Bay residents? Given that this pattern is also evident for the recycling policy, it is unlikely that it is linked to differences in perceived personal relevance. Rather, I tentatively suggest two other possible explanations for that. First, these patterns are resulted by the lower baseline of Haifa Bay residents' trust in EPM policies. Accordingly, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects), indicating a ceiling effect. In other words, symbols may be more effective in reducing citizens' distrust, then increasing their trust.[^footnote_ceiling_effect] A second explanation could be that EPM symbols attenuated the effect of Haifa Bay residents' prior beliefs on their judgements of the policy plans, an effect that can be related to motivated reasoning [@taber_2006; @christensen_2018; @baekgaard_2016]. It can be theorized that symbols attenuate motivated reasoning, when symbolic associations are not strongly related to identities of groups having strong prior positions on the policy or the organization (as in the case of the EPM symbols and the Haifa Bay local residents). On the other hand, when symbolic associations do activate such group identities, they are likely to exacerbate partisan motivated reasoning in that regard, and polarization [@teodoro_2018]. Certainly, we must be cautious about these tentative explanations, which may be tested by future studies.

[^footnote_ceiling_effect]: This explanation is supported by the comparison of the percentages of subjects who reported very low levels of trust in the policies. This comparison consistently shows a pattern of decreased marginal reduction. In the Haifa Bay, the percentage of those who reported a trust score of less than 2 in the air-pollution policy decreased from `r ((filter(SVIVA2_01_haifa,SYMBOL==0,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==0)%>%nrow())*100)%>%round(1)`% in the no symbols group to `r ((filter(SVIVA2_01_haifa,SYMBOL==2,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==2)%>%nrow())*100)%>%round(1)`% in the real symbols. Among the center subjects, only `r ((filter(SVIVA2_01_center,SYMBOL==0,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==0)%>%nrow())*100)%>%round(1)`% of the no symbols group reported a trust score lower than 2 in the first place, and this percentage dropped to `r ((filter(SVIVA2_01_center,SYMBOL==2,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==2)%>%nrow())*100)%>%round(1)`% among the real symbols group. With regard to the recycling policy, the percentage decreased from `r ((filter(SVIVA2_01_haifa,SYMBOL==0,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==0)%>%nrow())*100)%>%round(1)`% to `r ((filter(SVIVA2_01_haifa,SYMBOL==2,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==2)%>%nrow())*100)%>%round(1)`% in the Haifa Bay and from `r ((filter(SVIVA2_01_center,SYMBOL==0,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==0)%>%nrow())*100)%>%round(1)`% to `r ((filter(SVIVA2_01_center,SYMBOL==2,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==2)%>%nrow())*100)%>%round(1)`% in the center.

Still, several limitations of this study should be acknowledged. The study focuses on one policy domain, and on a specific, local issue. Future studies may put these findings to further test by exploring cases of air-pollution in other locations, as well as by exploring cases of other environmental policies, and other domains. Additionally, I cannot entirely rule out the possibility that the differences in personal relevance between the two geographic areas are confounded with other differences associated with people's beliefs about the subject and the agency. Future studies may conduct more rigorous investigations through natural experimental or quasi-experimental designs that could take advantage of particular cases of citizens' random allocation to varying levels of personal relevance. Finally, whereas this study includes an experimental manipulation for perceived personal relevance, I acknowledge the limitations of that test, namely its limited statistical power and the fact that the survey data does not confirm the effectiveness of the manipulation in altering the moderating variable. Therefore, future studies may explore different, possibly stronger experimental manipulations for perceived personal relevance.

The results of this study, if generalizable, have significant implications. They suggest that citizens are susceptible to persuasion by government branding and symbolic communication, even with regard to salient, consested policy issues that have significant consequences for their own lives. Accordingly, these findings suggest that the boundaries of persuasion via branding are much wider than expected. This entails that public organizations can affect citizens in a broad array of policy domains and contexts by investing in symbolic communications. Alongside the benefits of this instrument in mitigating citizens' distrust, it also has the potency to yield a distorted, emotive, overoptimistic view of government organizations and their policies, which may undermine basic democratic accountability mechanisms. For good or for ill, citizens are affected by the emotional connections and associations of symbols.
